---
title: "Session 9 - Single Cell RNA seq analysis with Seurat"
author: "I. Ali, M. Wells"
date: "`r Sys.Date()`"
output: word_document
---

# Introduction to Single Cell RNA Seq and Seurat

Single Cell RNA seq is an increasingly common analytical tool used to understand biological systems. For this session we are copying the Seurat vignette so that we can adapt it for our session. To get the original source for the tutorial [follow this link](https://satijalab.org/seurat/articles/pbmc3k_tutorial.html)

By the end of this session you should be able to.

* Recognize key functions in the Seurat package
* Recognize Seurat output plots
** Variable Feature Plots
** Elbow Plot
** JackStraw Plot
* Save Seurat output objects

# Seurat Vignette

In this tutorial, we will be analyzing the a dataset of Peripheral Blood Mononuclear Cells (PBMC) freely available from 10X Genomics. There are 2,700 single cells that were sequenced on the Illumina NextSeq 500. The data for this vignette has already been downloaded into our datasets folder.

# Setup

## Load Libraries

```{r setup, echo = FALSE}
rm(list=ls())
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse) # data formatting 
library(dplyr)
library(Seurat)
library(patchwork)

```

## Set up the Seurat Object (Data Import)

We start by reading in the data. The Read10X() function reads in the output of the cellranger pipeline from 10X, returning a unique molecular identified (UMI) count matrix. The values in this matrix represent the number of molecules for each feature (i.e. gene; row) that are detected in each cell (column).

We next use the count matrix to create a Seurat object using the `CreateSeuratObject()` command. The object serves as a container that contains both data (like the count matrix) and analysis (like PCA, or clustering results) for a single-cell dataset. For a technical discussion of the Seurat object structure, check out our GitHub Wiki. For example, the count matrix is stored in pbmc[["RNA"]]@counts.

```{r dimport, echo = F}

# Load the PBMC dataset from the output of the cellranger pipeline.
# notice that the file path goes to the directory, not a specific file
pbmc.data <- Read10X(data.dir = "Datasets/filtered_gene_bc_matrices/hg19/")

# Initialize the Seurat object with the raw (non-normalized data).
pbmc <- CreateSeuratObject(counts = pbmc.data, project = "pbmc3k", min.cells = 3, min.features = 200)
pbmc

```

# Standard pre-processing workflow

The steps below encompass the standard pre-processing workflow for scRNA-seq data in Seurat. These represent the selection and filtration of cells based on QC metrics, data normalization and scaling, and the detection of highly variable features.

## QC and selecting cells for further analysis

Seurat allows you to easily explore QC metrics and filter cells based on any user-defined criteria. A few QC metrics commonly used by the community include

* The number of unique genes detected in each cell.
** Low-quality cells or empty droplets will often have very few genes
** Cell doublets or multiplets may exhibit an aberrantly high gene count
* Similarly, the total number of molecules detected within a cell (correlates strongly with unique genes)
* The percentage of reads that map to the mitochondrial genome
** Low-quality / dying cells often exhibit extensive mitochondrial contamination
** We calculate mitochondrial QC metrics with the PercentageFeatureSet() function, which calculates the percentage of counts originating from a set of features
** We use the set of all genes starting with MT- as a set of mitochondrial genes

The number of unique genes and total molecules are automatically calculated during CreateSeuratObject(). You can find them stored in the object meta data. To view see the code below.

```{r QCsteps, echo = T}

# The [[ operator can add columns to object metadata. This is a great place to stash QC stats
pbmc[["percent.mt"]] <- PercentageFeatureSet(pbmc, pattern = "^MT-")

# Show QC metrics for the first 5 cells
head(pbmc@meta.data, 5)

```

Let's visualize our QC metrics with a violin plot, and use these to filter cells.

* We filter cells that have unique feature counts over 2,500 or less than 200
* We filter cells that have >5% mitochondrial counts

```{r QCviz, echo = T, fig.width="6", fig.height="5"}
# Visualize QC metrics as a violin plot
VlnPlot(pbmc, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
```


```{r QCscatter, echo=, fig.height=5, fig.width=7}

# FeatureScatter is typically used to visualize feature-feature relationships, but can be used
# for anything calculated by the object, i.e. columns in object metadata, PC scores etc.

plot1 <- FeatureScatter(pbmc, feature1 = "nCount_RNA", feature2 = "percent.mt")
plot2 <- FeatureScatter(pbmc, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
plot1 + plot2

# using subset, you complete the filtering step
pbmc <- subset(pbmc, subset = nFeature_RNA > 200 & nFeature_RNA < 2500 & percent.mt < 5)

```

# Normalizing the Data

After removing unwanted cells from the dataset, the next step is to normalize the data. By default, we employ a global-scaling normalization method “LogNormalize” that normalizes the feature expression measurements for each cell by the total expression, multiplies this by a scale factor (10,000 by default), and log-transforms the result. Normalized values are stored in `pbmc[["RNA"]]@data`.

```{r normalize, echo = T}

pbmc <- NormalizeData(pbmc, normalization.method = "LogNormalize", scale.factor = 10000)

# note: the Seurat vignette often shows the default arguments, in this case:
## normalization.method == "LogNormalize"
## scale.factor = 10000

# the command below would work in the same way
pbmc <- NormalizeData(pbmc)

```

# Identification of highly variable features (feature selection)

We next calculate a subset of features that exhibit high cell-to-cell variation in the dataset (i.e, they are highly expressed in some cells, and lowly expressed in others). We and others have found that focusing on these genes in downstream analysis helps to highlight biological signal in single-cell datasets.

Our procedure in Seurat is described in detail here, and improves on previous versions by directly modeling the mean-variance relationship inherent in single-cell data, and is implemented in the `FindVariableFeatures()` function. By default, we return 2,000 features per dataset. These will be used in downstream analysis, like PCA.

```{r id.var.featrues, echo=, fig.height=3, fig.width=7}

pbmc <- FindVariableFeatures(pbmc, selection.method = "vst", nfeatures = 2000)

# Identify the 10 most highly variable genes
top10 <- head(VariableFeatures(pbmc), 10)

# plot variable features with and without labels
plot1 <- VariableFeaturePlot(pbmc)
plot2 <- LabelPoints(plot = plot1, points = top10, repel = TRUE)
plot1 + plot2

```

# Scaling the Data

Next, we apply a linear transformation (‘scaling’) that is a standard pre-processing step prior to dimensional reduction techniques like PCA. The `ScaleData()` function:

* Shifts the expression of each gene, so that the mean expression across cells is 0
* Scales the expression of each gene, so that the variance across cells is 1
** This step gives equal weight in downstream analyses, so that highly-expressed genes do not dominate
* The results of this are stored in `pbmc[["RNA"]]@scale.data`

```{r scaling, echo = T}

all.genes <- rownames(pbmc)
pbmc <- ScaleData(pbmc, features = all.genes)

```

Scaling is an essential step in the Seurat workflow, but only on genes that will be used as input to PCA. Including all genes may take a long time. Therefore, the default in ScaleData() is only to perform scaling on the previously identified variable features (2,000 by default). To do this, omit the `features` argument in the previous function call, like we have done below. 

```{r scalingdata, echo = T}

# Before running this line, notice that the pbmc Seurat object is a 357MB object
pbmc <- ScaleData(pbmc)

# Scaling the data on only the previously identified variable features will reduce the size of the pbmc Seurat object to an 102.9MB object. 
```

Your PCA and clustering results will be unaffected. However, Seurat heatmaps (produced as shown below with DoHeatmap()) require genes in the heatmap to be scaled, to make sure highly-expressed genes don’t dominate the heatmap. To make sure we don’t leave any genes out of the heatmap later, we are scaling all genes in this tutorial.

# Perform linear dimensional reduction

Next we perform PCA on the scaled data. By default, only the previously determined variable features are used as input, but can be defined using `features` argument if you wish to choose a different subset.

```{r PCA, echo = T}

pbmc <- RunPCA(pbmc, features = VariableFeatures(object = pbmc))

```

Seurat provides several useful ways of visualizing both cells and features that define the PCA, including `VizDimReduction()`, `DimPlot()`, and `DimHeatmap()`

```{r PVAviz, echo = T}

# Examine and visualize PCA results a few different ways
print(pbmc[["pca"]], dims = 1:5, nfeatures = 5)

VizDimLoadings(pbmc, dims = 1:2, reduction = "pca")

DimPlot(pbmc, reduction = "pca")

```

In particular `DimHeatmap()` allows for easy exploration of the primary sources of heterogeneity in a dataset, and can be useful when trying to decide which PCs to include for further downstream analyses. Both cells and features are ordered according to their PCA scores. Setting cells to a number plots the ‘extreme’ cells on both ends of the spectrum, which dramatically speeds plotting for large datasets. Though clearly a supervised analysis, we find this to be a valuable tool for exploring correlated feature sets.

```{r Heatmap, echo=, fig.height=5, fig.width=7}

DimHeatmap(pbmc, dims = 1, cells = 500, balanced = TRUE)

DimHeatmap(pbmc, dims = 1:6, cells = 500, balanced = TRUE)
# You can continue to look at the other dimensions by modifying the dims argument.

```

To overcome the extensive technical noise in any single feature for scRNA-seq data, Seurat clusters cells based on their PCA scores, with each PC essentially representing a ‘metafeature’ that combines information across a correlated feature set. The top principal components therefore represent a robust compression of the dataset. However, how many components should we choose to include? 10? 20? 100?

In [Macosko et al](http://www.cell.com/abstract/S0092-8674(15)00549-8), we implemented a resampling test inspired by the JackStraw procedure. We randomly permute a subset of the data (1% by default) and rerun PCA, constructing a ‘null distribution’ of feature scores, and repeat this procedure. We identify ‘significant’ PCs as those who have a strong enrichment of low p-value features.

The `JackStrawPlot()` function provides a visualization tool for comparing the distribution of p-values for each PC with a uniform distribution (dashed line). ‘Significant’ PCs will show a strong enrichment of features with low p-values (solid curve above the dashed line). In this case it appears that there is a sharp drop-off in significance after the first 10-12 PCs.

```{r data.dim, echo = T}
# NOTE: This process can take a long time for big datasets, comment out for expediency. More
# approximate techniques such as those implemented in ElbowPlot() can be used to reduce
# computation time
pbmc <- JackStraw(pbmc, num.replicate = 100)
pbmc <- ScoreJackStraw(pbmc, dims = 1:20)

JackStrawPlot(pbmc, dims = 1:15)

```

An alternative heuristic method generates an ‘Elbow plot’: a ranking of principle components based on the percentage of variance explained by each one (`ElbowPlot()` function). In this example, we can observe an ‘elbow’ around PC9-10, suggesting that the majority of true signal is captured in the first 10 PCs.

```{r elbowplot, echo = T}

ElbowPlot(pbmc)

```

Identifying the true dimensionality of a dataset – can be challenging/uncertain for the user. We therefore suggest these three approaches to consider. The first is more supervised, exploring PCs to determine relevant sources of heterogeneity, and could be used in conjunction with GSEA for example. The second implements a statistical test based on a random null model, but is time-consuming for large datasets, and may not return a clear PC cutoff. The third is a heuristic that is commonly used, and can be calculated instantly. In this example, all three approaches yielded similar results, but we might have been justified in choosing anything between PC 7-12 as a cutoff.

We chose 10 here, but encourage users to consider the following:

* Dendritic cell and NK aficionados may recognize that genes strongly associated with PCs 12 and 13 define rare immune subsets (i.e. MZB1 is a marker for plasmacytoid DCs). However, these groups are so rare, they are difficult to distinguish from background noise for a dataset of this size without prior knowledge.
* We encourage users to repeat downstream analyses with a different number of PCs (10, 15, or even 50!). As you will observe, the results often do not differ dramatically.
* We advise users to err on the higher side when choosing this parameter. For example, performing downstream analyses with only 5 PCs does significantly and adversely affect results.

# Cluster the cells

Seurat v3 applies a graph-based clustering approach, building upon initial strategies in ([Macosko et al](http://www.cell.com/abstract/S0092-8674(15)00549-8)). Importantly, the distance metric which drives the clustering analysis (based on previously identified PCs) remains the same. However, our approach to partitioning the cellular distance matrix into clusters has dramatically improved. Our approach was heavily inspired by recent manuscripts which applied graph-based clustering approaches to scRNA-seq data [SNN-Cliq, Xu and Su, Bioinformatics, 2015](http://bioinformatics.oxfordjournals.org/content/early/2015/02/10/bioinformatics.btv088.abstract) and CyTOF data [PhenoGraph, Levine et al., Cell, 2015](http://www.ncbi.nlm.nih.gov/pubmed/26095251). Briefly, these methods embed cells in a graph structure - for example a K-nearest neighbor (KNN) graph, with edges drawn between cells with similar feature expression patterns, and then attempt to partition this graph into highly interconnected ‘quasi-cliques’ or ‘communities’.

As in PhenoGraph, we first construct a KNN graph based on the euclidean distance in PCA space, and refine the edge weights between any two cells based on the shared overlap in their local neighborhoods (Jaccard similarity). This step is performed using the `FindNeighbors()` function, and takes as input the previously defined dimensionality of the dataset (first 10 PCs).

To cluster the cells, we next apply modularity optimization techniques such as the Louvain algorithm (default) or SLM [SLM, Blondel et al., Journal of Statistical Mechanics](http://dx.doi.org/10.1088/1742-5468/2008/10/P10008), to iteratively group cells together, with the goal of optimizing the standard modularity function. The `FindClusters()` function implements this procedure, and contains a resolution parameter that sets the ‘granularity’ of the downstream clustering, with increased values leading to a greater number of clusters. We find that setting this parameter between 0.4-1.2 typically returns good results for single-cell datasets of around 3K cells. Optimal resolution often increases for larger datasets. The clusters can be found using the `Idents()` function.

```{r clustering}

pbmc <- FindNeighbors(pbmc, dims = 1:10)
pbmc <- FindClusters(pbmc, resolution = 0.5)

# Look at cluster IDs of the first 5 cells
head(Idents(pbmc), 5)

```

# Run non-linear dimensional reduction (UMAP/tSNE)

Seurat offers several non-linear dimensional reduction techniques, such as tSNE and UMAP, to visualize and explore these datasets. The goal of these algorithms is to learn the underlying manifold of the data in order to place similar cells together in low-dimensional space. Cells within the graph-based clusters determined above should co-localize on these dimension reduction plots. As input to the UMAP and tSNE, we suggest using the same PCs as input to the clustering analysis.

```{r UMAP, echo = T}

# If you haven't installed UMAP, you can do so via reticulate::py_install(packages =
# 'umap-learn')
pbmc <- RunUMAP(pbmc, dims = 1:10)

# note that you can set `label = TRUE` or use the LabelClusters function to help label
# individual clusters
DimPlot(pbmc, reduction = "umap")


```

You can save the object at this point so that it can easily be loaded back in without having to rerun the computationally intensive steps performed above, or easily shared with collaborators.

```{r save, echo = T}

saveRDS(pbmc, file = "Output/pbmc_tutorial.rds")

```

# Finding differentially expressed features (cluster biomarkers)

Seurat can help you find markers that define clusters via differential expression. By default, it identifies positive and negative markers of a single cluster (specified in `ident.1`), compared to all other cells.  `FindAllMarkers()` automates this process for all clusters, but you can also test groups of clusters vs. each other, or against all cells.

The `min.pct` argument requires a feature to be detected at a minimum percentage in either of the two groups of cells, and the thresh.test argument requires a feature to be differentially expressed (on average) by some amount between the two groups. You can set both of these to 0, but with a dramatic increase in time - since this will test a large number of features that are unlikely to be highly discriminatory. As another option to speed up these computations, `max.cells.per.ident` can be set. This will downsample each identity class to have no more cells than whatever this is set to. While there is generally going to be a loss in power, the speed increases can be significant and the most highly differentially expressed features will likely still rise to the top.

```{r markers1, fig.height=8, fig.width=15}
# find all markers of cluster 2
cluster2.markers <- FindMarkers(pbmc, ident.1 = 2, min.pct = 0.25)
head(cluster2.markers, n = 5)
# find all markers distinguishing cluster 5 from clusters 0 and 3
cluster5.markers <- FindMarkers(pbmc, ident.1 = 5, ident.2 = c(0, 3), min.pct = 0.25)
head(cluster5.markers, n = 5)
# find markers for every cluster compared to all remaining cells, report only the positive ones
pbmc.markers <- FindAllMarkers(pbmc, only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25)
pbmc.markers %>% group_by(cluster) %>% slice_max(n = 2, order_by = avg_log2FC)
```

Seurat has several tests for differential expression which can be set with the test.use parameter (see our [DE vignette](de_vignette.html) for details). For example, the ROC test returns the 'classification power' for any individual marker (ranging from 0 - random, to 1 - perfect).

```{r markersroc, fig.height=8, fig.width=15}
cluster0.markers <- FindMarkers(pbmc, ident.1 = 0, logfc.threshold = 0.25, test.use = "roc", only.pos = TRUE)
```

We include several tools for visualizing marker expression. `VlnPlot()` (shows expression probability distributions across clusters), and `FeaturePlot()` (visualizes feature expression on a tSNE or PCA plot) are our most commonly used visualizations. We also suggest exploring `RidgePlot()`, `CellScatter()`, and `DotPlot()` as additional methods to view your dataset.

```{r markerplots, fig.height=10, fig.width=15}
VlnPlot(pbmc, features = c("MS4A1", "CD79A"))
# you can plot raw counts as well
VlnPlot(pbmc, features = c("NKG7", "PF4"), slot = 'counts', log = TRUE)
FeaturePlot(pbmc, features = c("MS4A1", "GNLY", "CD3E", "CD14", "FCER1A", "FCGR3A", "LYZ", "PPBP", "CD8A"))
```

`DoHeatmap()` generates an expression heatmap for given cells and features. In this case, we are plotting the top 20 markers (or all markers if less than 20) for each cluster.

```{r clusterHeatmap, fig.height=8, fig.width=15}
pbmc.markers %>% group_by(cluster) %>% top_n(n = 10, wt = avg_log2FC) -> top10
DoHeatmap(pbmc, features = top10$gene) + NoLegend()
```


***

# Assigning cell type identity to clusters

Fortunately in the case of this dataset, we can use canonical markers to easily match the unbiased clustering to known cell types:

Cluster ID | Markers       | Cell Type
-----------|---------------|----------
0          | IL7R, CCR7    | Naive CD4+ T 
1          | CD14, LYZ     | CD14+ Mono
2          | IL7R, S100A4  | Memory CD4+ 
3          | MS4A1         | B 
4          | CD8A          | CD8+ T 
5          | FCGR3A, MS4A7 | FCGR3A+ Mono
6          | GNLY, NKG7    | NK 
7          | FCER1A, CST3  | DC
8          | PPBP          | Platelet


```{r labelplot, fig.height=5, fig.width=9}
new.cluster.ids <- c("Naive CD4 T", "CD14+ Mono", "Memory CD4 T", "B", "CD8 T", "FCGR3A+ Mono", "NK", "DC", "Platelet")
names(new.cluster.ids) <- levels(pbmc)
pbmc <- RenameIdents(pbmc, new.cluster.ids)
DimPlot(pbmc, reduction = 'umap', label = TRUE, pt.size = 0.5) + NoLegend()
```


```{r save.img, include=FALSE}
library(ggplot2)
plot <- DimPlot(pbmc, reduction = "umap", label = TRUE, label.size = 4.5) + xlab("UMAP 1") + ylab("UMAP 2") + 
  theme(axis.title = element_text(size = 18), legend.text = element_text(size = 18)) + 
  guides(colour = guide_legend(override.aes = list(size = 10)))
ggsave(filename = "Output/pbmc3k_umap.jpg", height = 7, width = 12, plot = plot, quality = 50)
```


```{r save.rds, eval=FALSE}
saveRDS(pbmc, file = "Output/pbmc3k_final.rds")
```


```{r save2, include = FALSE}
saveRDS(pbmc, file = "Datasets/pbmc3k_final.rds")
```

# Conclusion

Over the sessions in this course we have focused on common statistical and analytical strategies used in biological and biomedical research. Our sessions have been split between the theoretical foundation behind the statistical tests used in research, and the practical skills to apply these tools using R coding. We have covered t-testing, ANOVA, correlation analysis, linear regression, PCA and clustering. In this session we have done a deep dive into single-cell RNA seq, and applied several of the visualization tools and multidimensional data analysis strategies we have convered in previous sessions. We used the Seruat vignette to run a single cell RNA seq analysis on the output of the CellRanger pipeline. We performed several quality control steps, and used multiple visualization strategies to understand the output of this analysis. 

For our second and final assignment we are going to get a chance to apply the things we have learned in this vignette to a new data set. We will break the assignment up in to tasks, similar to Assignment 1 and you will be able to work with your peers to complete the assignment.